---
title: "Stat428_FinalProj"
author: "Vetrie Senthilkumar, Chris Zhu"
date: "November 26, 2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Method 
To test the viability of using Monte Carlo Simulations to predict stock prices, we decided to simulate the price of a particularly volatile stock: Tesla. Compared to other stocks, Tesla's behavior is more difficult to predict since it rarely depicts a clear trend. This indicates that Tesla's stock fluctuations might be best captured by a random walk based model. 

## Data 
We accessed historical stock data through an R library called quantmod. Quantmod provided details about the opening, closing, high, and low price of Tesla's stock on a daily basis from June 6th, 2010 to present day.  Quantmod also contains a function that computes the daily returns. We subsetted the daily returns data based on training period dates to obtain the input to the model. Similarly, we subsetted the closing price data based on the test period dates to obtain the actual stock prices to compare our simulations against. 


## Model 

#### Applying GBM to Stock Prices
We used a MCMC process based on Geometric Brownian Motion (GBM) to simulate Tesla's stock price. GBM is continuous-time stochastic process often used to predict the movement of time series variables such as financial asset prices. GBM solves the following stochastic differential equation: 

$dS_t = \mu S_t dt + \sigma S_t dB_t$

Solving this equation produces the following formula for GBM:

$S_{t+1} = S_te^{(\mu - \frac{\sigma^2}{2})t + \sigma B_t}$  

For the purposes of our project, $\mu$ represents the mean of the daily returns and $\sigma$ represents the standard deviation of the daily returns. $S_0$ is the most recent closing price of the stock. $B_t$ is a standard normal random variable and $t$ is time period between consecutive simulated stock price values. In our simulation, we assume a day has elapsed between successive draws. Thus, $t = 1$ and $t$ can be omitted entirely from our formula for GBM.

The exponent in the GBM consists of two components: the drift and the shock. The drift is given by $(\mu - \frac{\sigma^2}{2})t$ while the shock is given by 
$\sigma B_t$. The drift takes into consideration the long term direction of the stock's price while the shock accounts for volatility and randomness. The shock induces randomness into the model through the standard normal random variable $B_t$. Thus, GBM incorporates two seemingly contradictory behaviors observed in stock prices: long term trends and short term price fluctuations. 

It should be mentioned that GBM relies on the following assumptions:

1. Stock prices follow a Markov process which implies that we only need the current price of the stock to predict future prices (This can be seen in the formula for GBM. $S_{t+1}$ is only depedent on $S_t$, not on any previous prices) 
2. The returns of stocks are normally distributed 
3. The price levels of stocks are log-normally distributed 

#### Weighted Means and Standard Deviation

One prevalent issue with GBM is that all historical returns of stock are weighted equally, but intuitively we expect the current trend to have a larger impact on the future stock price. Generally speaking, the returns of a stock are better reflected by recent returns than returns that occurred years ago. To reflect this behavior, we wanted to create a weighted version of our original model that places greater importance on recent returns. In essence, this process would be similar to a type of importance sampling. In the new weighted model, we replaced $\mu$ and $\sigma$ with the following:

$\hat\mu = \frac{\sum_{i=1}^Nw_ix_i}{\sum_{i=1}^Nw_i}$

$\hat\sigma = \sqrt{\frac{\sum_{i=1}^Nw_i}{(\sum_{i=1}^Nw_i)^2 - (\sum_{i=1}^Nw_i^2)}\sum_{i=1}^Nw_i(x_i-\mu)^2}$

The vector of weights consists of $w_i = (1 - \alpha)^{N - i}$ where $\alpha$ is a constant. Our simulations performed best when $\alpha$ was set to 0.01. 

#### Computing Confidence Intervals 

As mentioned previously, one of the assumptions of GBM is that the price levels of the stock are log normally distributed. Therefore, we felt that it would be better idea to construct a standard normal confidence interval using the log of the predicted stock prices instead. To do so, we apply a logarithmic transformation on the predicted stock prices generated by the simulation and compute the corresponding confidence interval. Afterward, we apply an exponential transformation to the upper and lower bound of the confidence interval to obtain a confidence interval for the regular price of the stock.


## Algorithm 

Since most of our model was constructed as a single function, we decided to provide code to facilitate an understanding of our implementation.

```{r}
runMCSims = function(num_sim, total_days, stock_data, alpha) {
  returns = dailyReturn(stock_data)
 
  # Check whether to use weighted or unweighted parameters 
  if (alpha > 0) {
    weighted_params = computeWeightedParams(alpha, returns)
    mu = weighted_params[1]
    sigma = weighted_params[2]
  } else {
    mu = mean(returns)
    sigma = sd(returns)
  }
  
  sim_mat = matrix(0, nrow = num_sim, ncol = total_days + 1)
  sim_mat[, 1] = tail(Cl(stock_data), 1)[[1]] # All sims start off with most recent closing price
  
  drift = (mu - sigma^2 / 2)

  for (k in 1:num_sim) {
    for (i in 1:total_days) {
      shock = sigma * rnorm(1)
      sim_mat[k, i + 1] = sim_mat[k, i] * exp(drift + shock)
    }
  } 
  
  return (sim_mat)
} 
```

The function representing our model requires four inputs: the number of simulations that should be run (num_sim), the number of days each simulation should try to predict stock prices for (total_days), the ticker symbol of the stock (ticker), and the weight (alpha). When alpha is 0, the regular, unweighted model is used. 

1. First use the relevant stock data to compute the daily returns 

2. Determine whether to use weighted or unweighted parameters. If weighted parameters are required, calculate the correct parameters using the helper function computeWeightedParams(). This helper function computes the formulas discussed in the Weighted Mean and Standard Deviation section. Otherwise, set mu and sigma as the mean and standard deviation of the historical daily returns of the stock

3. We create an empty matrix (sim_mat) to store the results of our simulations. Each row in the matrix represents a single simulation and contains the generated stock prices. The rows can also be thought of as GBM Markov Chains. Each row has (total_days + 1) entries, or an entry for each of the days we want to simulate the future price. The very first, additional entry is set to the most recent closing price. We include this information because our Markov Chain needs a starting point to proceed from. This entry should be omitted for all analyses. 

4. The drift is computed using mu and sigma. Note that that drift is a constant quantity and is independent of where we are in the GBM Markov Chain. Thus, drift can be calculated prior to any sort of iterative method used to generate the chain.

5. The outer loop iterates through the number of simulations that need to be run. The inner loop iterates through the number of days we are simulating prices for. For each iteration of the inner loop, the stock price at the next day is calculated using the current stock price, the shock, and the drift. The function rnorm() is used to generate a standard normal random variable that adds an element of randomness to the shock. The next day's stock price is then stored into the matrix. Thus, we are filling in sim_mat row by row.   

6. The first column of sim_mat containing init_price is removed and the matrix is returned as output 


# Results and Code
```{r}
library(quantmod)
library(ggplot2)
library(rbenchmark)
library(magrittr)
```

```{r}
# Confidence interval code 
ci = function(sim_mat, conf_level) {
  final_prices = sim_mat[, ncol(sim_mat)]
  
  mu = mean(log(final_prices))
  sigma = sd(log(final_prices))
  
  Z = qnorm((1 - conf_level)/2) 
  
  ln_final_price_CI = c(mu + Z*sigma, mu - Z*sigma)
  final_price_CI = exp(ln_final_price_CI)
  
  return (final_price_CI)
}
```


```{r}
# Function to compute weighted parameters 
computeWeightedParams = function(alpha, returns) {
  weights = numeric(length(returns))
  
  for (i in 1:length(weights)) {
    weights[i] = (1 + alpha)^i
  }
  
  diff = (returns - mean(returns))^2
  
  mu = sum(weights * returns) / sum(weights)
  sigma = sqrt((sum(weights) * sum(diff * weights)) / (sum(weights)^2 - sum(weights^2)))
  
  return (c(mu, sigma))
}

runMCSims = function(num_sim, total_days, stock_data, alpha) {
  returns = dailyReturn(stock_data)
 
  # Check whether to use weighted or unweighted parameters 
  if (alpha > 0) {
    weighted_params = computeWeightedParams(alpha, returns)
    mu = weighted_params[1]
    sigma = weighted_params[2]
  } else {
    mu = mean(returns)
    sigma = sd(returns)
  }
  
  sim_mat = matrix(0, nrow = num_sim, ncol = total_days + 1)
  sim_mat[, 1] = tail(Cl(stock_data), 1)[[1]] # All sims start off with most recent closing price
  
  drift = (mu - sigma^2 / 2)

  for (k in 1:num_sim) {
    for (i in 1:total_days) {
      shock = sigma * rnorm(1)
      sim_mat[k, i + 1] = sim_mat[k, i] * exp(drift + shock)
    }
  } 
  
  return (sim_mat)
}
```


```{r}
summarizeSimResults = function(ticker) {
  periods = c()
  closing_prices = c()
  
  lb = c()
  ub = c()
  contains = c()
  
  lb_w = c()
  ub_w = c()
  contains_w = c()
  
  # Load stock data
  data = getSymbols(ticker, auto.assign = FALSE)
  returns = dailyReturn(data)
  
  # Dates defining month, quarter, year, and three years
  PERIOD_DATA = list(list('Month', '2019-10-31', '/2019-10-31', '2019-11-01/'), list('Quarter', '2019-09-30', "/2019-09-30", '2019-10-01/'), list('Year', '2018-12-31', '/2018-12-31', '2019-01-01/'), list('3 Years', '2015-12-31', '/2015-12-31', '2016-01-01/'))
  
  for (i in 1:length(PERIOD_DATA)) {
    cur_period = as.character(PERIOD_DATA[[i]][1])
    start_date = as.character(PERIOD_DATA[[i]][2])
    start_date_slash = as.character(PERIOD_DATA[[i]][3])
    end_date = as.character(PERIOD_DATA[[i]][4])
    
    closing_price = Cl(data)[start_date][[1]]
    train_returns = returns[start_date_slash]
    test_prices = Cl(data)[end_date]
    actual_close = test_prices[length(test_prices)]
    
    # Run unweighted and weighted simulations
    sim_mat = runMCSims(25, length(test_prices), data, 0)
    sim_mat_w = runMCSims(25, length(test_prices), data, .01)
    
    # Compute confidence intervals using results of simulations
    conf_int = ci(sim_mat, .9)
    price_in_ci = actual_close >= conf_int[1] && actual_close <= conf_int[2]
    
    conf_int_w = ci(sim_mat_w, .9)
    price_in_ci_w = actual_close >= conf_int_w[1] && actual_close <= conf_int_w[2]
    
    periods = c(periods, cur_period)
    closing_prices = c(closing_prices, actual_close)
    
    # Check if confidence intervals contain true stock price 
    lb = c(lb, conf_int[1])
    ub = c(ub, conf_int[2])
    contains = c(contains, price_in_ci)
    
    lb_w = c(lb_w, conf_int_w[1])
    ub_w = c(ub_w, conf_int_w[2])
    contains_w = c(contains_w, price_in_ci_w)
    
    # Creating y limits for graphing purposes 
    ylim_nonweighted = c(min(test_prices, sim_mat), max(test_prices, sim_mat))
    ylim_weighted = c(min(test_prices, sim_mat_w), max(test_prices, sim_mat_w))
                         
    # Plot stock prices generated using nonweighted model
    matplot(t(sim_mat), type = 'l', main = paste0(ticker, ' Share Prices (Nonweighted)'), ylab = 'Price ($)', xlab = 'Day', ylim = ylim_nonweighted)
    matlines(as.numeric(test_prices), lwd = 3)
    
    # Plot stock prices generated using weighted model
    matplot(t(sim_mat_w), type = 'l', main = paste0(ticker, ' Share Prices (Weighted)'), ylab = 'Price ($)', xlab = 'Day', ylim = ylim_weighted)
    matlines(as.numeric(test_prices), lwd = 3)
  }
  
  # Summarize confidence intervals in a table 
  data_summary = data.frame(Period = periods, closing_prices, lb, ub, contains)
  data_summary_w = data.frame(Period = periods, closing_prices, lb_w, ub_w, contains_w)
  colnames(data_summary) = c('Period', 'Closing Price', 'Lower Bound', 'Upper Bound', 'CI Contains')
  colnames(data_summary_w) = c('Period', 'Closing Price', 'Lower Bound', 'Upper Bound', 'CI Contains')
  
  print(data_summary)
  print(data_summary_w)
}
```

#TSLA
```{r}
summarizeSimResults('TSLA')
```

# AMZN
```{r}
summarizeSimResults('AMZN')
```

# JNJ
```{r}
summarizeSimResults('JNJ')
```
